RVD2: An ultra-sensitive variant detection model for low-depth targeted next-generation sequencing data

### 2013-12-06 doc/bioinformatics2013/RVD2.tex
updated results section with narrative of testing procedure.

The HCC1187 cell line is derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma. A matched normal cells were derived from lymphoblastoid cells from whole blood. Sequencing libraries were prepared according to the protocol described in the original technical report~\cite{}. The raw FASTQ read files were aligned to hg19 igenome~\cite{} using the Isaac aligner to generate BAM files~\citep{raczy2013isaac}. We used samtools mpileup to generate pileup files using hg19 igenome~\cite{} as reference sequence~\citep{Navin:2010gu}. The HCC1187 dataset is a well-recognized baseline dataset from Illumina for evaluating sequence analysis algorithms~\citep{newman2013relative,howarth2011large,howarth2007array}.
% TODO:cite the technical report.
% TODO: check matched cells line origin
% double checked by Yuting, didn't find the matched cells line origin
% TODO: mention tumor purity here

Sequencing libraries were prepared from 500 ng of DNA using an early access version of Illuminas TruSeq DNA PCR-Free kit. The libraries were sequenced on four and eight lanes of a HiSeq 2000 flow cell for the normal and tumor, respectively, using 100 bp paired-end reads. Data, analyzed using the pre-release version of the Illumina Cancer Sequencing Workflow showed an average coverage of 40x for the normal and 90x for the tumor DNA with about 96\% coverage with 10 or more reads~\cite{}.
% TODO: write this in own words.
% TODO: add citation here.

% TODO: discuss alignment protocol. Done.
% TODO: rvd2 parameters. Done.
% TODO: Pubmed for peer-reviewed paper on HCC1187 sample. Done

@@

%TODO: This section doesn't make sense to me. Lets talk about what needs to be done here.
Considering HCC1187 sample is diploid, we did two tests based on posterior density using RVD2. First, we compared the posterior density $\tilde\mu_j^{\Delta}$ to find out if $\tilde\mu_j$ in case sample was significantly higher than that in control sample and vice verse.Threshold $\tau$ was set at $1\times10^{-3}$ to avoid error brought by computer limitation; confidence level $1-\alpha$ was set at 0.95. We did $\chi^2$ test afterwards to further remove false callings.  % The difference is constant at 1.74*10^{-4}
\begin{equation}\label{eqn:bayes_test_positive}
	\Pr( \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}} \geq \tau ) > 1-\alpha,
\end{equation}
\begin{equation}\label{eqn:bayes_test_negative}
	\Pr( \tilde{\mu}_j^{\text{control}} - \tilde{\mu}_j^{\text{case}} \geq \tau ) > 1-\alpha,
\end{equation}

Second, we identified the germline status and somatic status of HCC1187 sample by looking into posterior density of control sample and case sample independently. Basically, We computed the empirical posterior probability of $\tilde\mu_j$ across interval $[\tau_\text{l},\tau_\text{u}]$ and compare it with confidence level $1-\alpha$,
\begin{equation}\label{eqn:bayes_test_interval}
	\Pr( \tilde{\mu}_j \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha.
\end{equation}

Confidence level $1-\alpha$ was set at 0.8 due to our lower confidence on germline and somatic status.Posterior density interval for reference, heterzygous mutation and homozygous mutation were set as $[0.0,0.2],[0.2,0.8],[0.8.1]$, respectively.
% End of section that is confusing to me.

>>>>

The HCC1187 cell line is derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma. A matched normal cells were derived from lymphoblastoid cells from whole blood. Sequencing libraries were prepared according to the protocol described in the original technical report~\cite{}. The raw FASTQ read files were aligned to hg19 igenome~\cite{} using the Isaac aligner to generate BAM files~\citep{raczy2013isaac}. The aligned data had an average read depth of 40x for the normal sample and 90x for the tumor sample with about 96\% coverage with 10 or more reads. We used samtools mpileup to generate pileup files using hg19 igenome~\cite{} as reference sequence~\citep{Navin:2010gu}. The HCC1187 dataset is a well-recognized baseline dataset from Illumina for evaluating sequence analysis algorithms~\citep{newman2013relative,howarth2011large,howarth2007array}.
% TODO:cite the technical report.
% TODO: check matched cells line origin
% double checked by Yuting, didn't find the matched cells line origin
% TODO: mention tumor purity here

% TODO: rvd2 parameters. Done.

@@

We used RVD2 to identify germline and somatic mutations in the diploid HCC1187 sample. To identify germline mutations, we used the control data in the following hypothesis test $\Pr( \tilde{\mu}_j \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha$, where the intervals are: homozygous reference $[0, 0.5]$, heterozygous mutant $[0.25, 0.75]$, and homozygous mutant $[0.5, 1.0]$ and the size of the test is $\alpha=0.05$.

To identify somatic mutations, we considered scenarios when the case(tumor) error rate is lower than the control(germline) error rate (e.g. loss-of-heterozygosity) as well as scenarios when the case(tumor) error rate is higher than the control(germline) error rate (e.g. homozygous somatic mutation). The two hypothesis tests are then $\Pr( \tilde{\mu}_j^{\Delta} \geq \tau ) > 1-\alpha$ and $\Pr( \tilde{\mu}_j^{\Delta} \leq \tau ) > 1-\alpha$. The size of the test is $\alpha=0.05$.
%TODO: correct notation from \Pr to summation because we have samples, not distributions.
